Learning constraints in tabular data

### 2016-05-13 Paper/ECAI/tabular.tex

The question that we answer in this paper is: whether it is possible to discover or reconstruct structural constraints (relations, functions) in flat tabular spreadsheet data?
To answer this question we contribute a general-purpose algorithm and system for discovering such constraints. Furthermore, the algorithm accepts declarative specifications of the constraints and predicates of interest (the "bias") and covers both textual and numeric fields. It works in an unsupervised setting, as only the tabular data is provided and no examples need to be specified, and discovers column- and row-wise constraints.
We demonstrate the utility of our approach in an experimental evaluation.
Moreover, we sketch several application scenario's, such as autocompletion and error detection.
\\\\
This paper is organized as follows.
Section \ref{sec:formalization} introduces concepts relevant to our approach. Section \ref{sec:problem_statement} presents the problem statement. Section \ref{sec:approach} describes the approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications such as autocompletion and error detection. Section \ref{sec:related_work} present an overview of the related work. And Section \ref{sec:conclusions} provides conclusions.

@@

\subsection{Cells, vectors and groups}
\tias{perhaps we should not introduce cells, we reason at the level of vectors!}
A spreadsheet may conceptually consist of multiple tables, such as in Figure~\ref{fig:main_example}. We will describe a method of extracting tables from a big spreadsheet in Section~\ref{todo}, however this is not the core of our learning method. Our formalization reasons over tables directly.

Formally, a table is an $n \times m$ matrix. Each entry is called a \textit{cell}.
A cell has a {\em type}, which can be numeric or textual. We further distinguish numeric types in subtypes integer and float. We also consider \textit{None} as a special type when a cell is empty; \textit{None} is a subtype of all other types.
A set of cells is called \textit{type-consistent} iff all cells in the set are of the same type.
%Certain constraints, such as \textit{rank} or \textit{series}, make use of the numeric subtype by requiring its arguments to be integers.

@@

\subsection{Constraint templates}
Our method will extract the applicable constraints (including functions) in a set of tables, based on \textit{constraint templates}.
%
A \template is a triple \textit{(\CName, \CSignature, \CFunction)}:
%Let us elaborate on this:
\begin{itemize}
\item
\textit{\CName}  specifies the syntactic form of the constraint $c(A_1, ...,A_n)$, that is, the name of the constraint $c$ together
with $n$ arguments $A_i$. Thus a constraint $c$ is viewed as a relation or predicate of arity $n$ in first order logic. Note that a function $B=f(A_1,...,A_n)$ can be represented with the $n+1$-ary predicate $cf(B,A_1,...,A_n)$.
% the list of its variables $v_1,\dots,v_n$.  In ILP terminology, it is known as vocabulary.

\item the \textit{\CSignature} defines the properties that the arguments of the predicate must satisfy. That is, the required types of the arguments, together
with possible further restrictions, e.g., on their length. For example, the signature can restrict the types of the arguments to be groups of integers and require that the length of all arguments be equal. \\
In terms of logical and relational learning \cite{luc_book}, such a \CSignature is known as the {\em bias} of the learner, it specifies when the constraint is well-formed.

@@

\paragraph{Example}
Several example constraints are illustrated in Table 1.  For example, the constraint template \textit{rank} has
\begin{itemize}
\item \CName: \textit{$B$ = RANK($A$)}, where $B$ and $A$ are the arguments;
\item \CSignature: the two arguments $B$ and $A$ must be numeric groups and of size 1; furthermore, the group $B$ must be integer (the rank) and the length of groups $A$ and $B$ must be identical, i.e., $length(B) = length(A)$;
\item \CFunction: \textit{$B$ = RANK($A$)} whenever each value of $B$ is the rank of the corresponding value in $A$.
% is satisfied in the mapping $\{ v_x \mapsto g_x, v_y \mapsto g_y \}$ satisfies the constraint iff $g_x \in G_x, g_y \in G_y$ ($G_x \in \groups, G_y \in \groups$) and each value in $g_y$ is the rank of the corresponding value in $g_x$ (possibly with ties).
\end{itemize}

When looking into constraints like $\ecrank{B}{A}$, it is helpful to see the analogy with both first order logic (FOL) and with constraint satisfaction problems.
From a FOL perspective, the name of the constraint ($RANK$) is just the predicate, and the arguments $B$ and $A$ are the terms, which can be seen as either uninstantiated variables or as concrete values.

This also holds in our setting, where an instantiation of a variable corresponds to a concrete set of vectors (a subgroup). For example for $\ecrank{B}{A}$ and the spreadsheet in Figure~\ref{fig:main_example}, when we write $\ecrank{\range{T_1}{\rangeall}{8}}{\range{T_1}{\rangeall}{7}}$, then the value of $B$ is the $8$th vector in $T_1$: $B = \range{T_1}{\rangeall}{8} = \{[2,1,3,4]\}$ and the value of $A$ is the $7$th vector: $A = \range{T_1}{\rangeall}{7} = \{[1514, 1551, 691, 392]\}$.
\tias{not OK to use set because of column-wise sum? we must properly define it then...}

With this interpretation, we can speak about the signature and definition of a constraint template being satisfied. We say that a signature (definition) of a constraint template with $n$ arguments is satisfied by an assignment $(A_1, ..., A_n)$ if $Sig_c(A_1, ..., A_n)$ (respectively $Def_c(A_1, ..., A_n)$) is satisfied. Likewise, an assignment $(A_1, ..., A_n)$ satisfies a constraint template $s$ iff $|(A_1, ..., A_n)| = n_s \wedge Sig_c(A_1, ..., A_n) \wedge Def_c(A_1, ..., A_n)$.

@@

\sergey{example new template: vr = floor(v), float(v), int(vr), len(v)=len(vr), v,vr are single vectors?}

>>>>

The question that we answer in this paper is: is it possible to discover or reconstruct structural constraints (relations, functions) in flat tabular spreadsheet data?
To answer this question we contribute a general-purpose algorithm and the system, named \sname (pronounced as the word ``tackle''), for discovering such constraints. Furthermore, the algorithm accepts declarative specifications of the constraints and predicates of interest (the "bias") and covers both textual and numeric fields. It works in an unsupervised setting, as only the tabular data is provided and no examples need to be specified, and discovers column- and row-wise constraints.
We demonstrate the utility of our approach in an experimental evaluation.
Moreover, we sketch several application scenario's such as autocompletion and error detection.
\\\\
This paper is organized as follows.
Section \ref{sec:formalization} introduces concepts relevant to our approach. Section \ref{sec:problem_statement} presents the problem statement. Section \ref{sec:approach} describes the approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications. Section \ref{sec:related_work} presents an overview of the related work. Section \ref{sec:conclusions} provides conclusions.

@@

\subsection{Cells, vectors and groups}
\tias{perhaps we should not introduce cells, we reason at the level of vectors!}
A spreadsheet may conceptually consist of multiple tables, such as in Figure~\ref{fig:main_example}. We will describe a method of extracting tables from a big spreadsheet in Section~\ref{sec:approach}, however this is not the core of our learning method. Our formalization reasons over tables directly.

Formally, a table is an $n \times m$ matrix. Each entry is called a \textit{cell}.
A cell has a {\em type}, which can be numeric or textual. We further distinguish numeric types in subtypes: integer and float. We also consider \textit{None} as a special type when a cell is empty; \textit{None} is a subtype of all other types.
A set of cells is called \textit{type-consistent} iff all cells in the set are of the same type (\textit{None} is consistent with everything).
%Certain constraints, such as \textit{rank} or \textit{series}, make use of the numeric subtype by requiring its arguments to be integers.

@@

\subsection{Constraint templates}
Our method will extract the applicable constraints (including functions) in a set of tables, based on \textit{constraint templates}.
%
A \template is a triple \textit{(\CName, \CSignature, \CFunction)}:
%Let us elaborate on this:
\begin{itemize}
\item
\textit{\CName}  specifies the syntactic form of the constraint $c(G_1, ...,G_n)$, that is, the name of the constraint $c$ together
with $n$ arguments $G_i$. Thus a constraint $c$ is viewed as a relation or predicate of arity $n$ in first order logic. Note that a function $G_r=f(G_1,...,G_n)$ can be represented with the $n+1$-ary predicate $cf(G_r,G_1,...,G_n)$.
% the list of its variables $v_1,\dots,v_n$.  In ILP terminology, it is known as vocabulary.

\item the \textit{\CSignature} defines the properties that the arguments of the predicate must satisfy. That is, the required types of the arguments, together
with possible further restrictions, e.g., on their length. For example, the signature can restrict the types of the arguments to be groups of integers and require that the length of all arguments be equal. In terms of logical and relational learning \cite{luc_book}, such a \CSignature is known as the {\em bias} of the learner, it specifies when the constraint is well-formed.

@@

\paragraph{Example}
Several example constraints are illustrated in Table \ref{table:constraints}.  For example, the constraint template \textit{rank} has
\begin{itemize}
  \item \CName: \ecrank{G_r}{G_x}, where $G_r$ and $G_x$ are the arguments;
\item \CSignature: the two arguments $G_r$ and $G_x$ must be numeric groups and of size one; furthermore, the group $G_r$ must be integer (the rank) and the length of groups $G_x$ and $G_r$ must be identical, i.e., $\plength(G_r) = \plength(G_x)$;
\item \CFunction: \ecrank{G_r}{G_x}, whenever each value of $G_r$ is the rank of the corresponding value in $G_x$.
% is satisfied in the mapping $\{ v_x \mapsto g_x, v_y \mapsto g_y \}$ satisfies the constraint iff $g_x \in G_x, g_y \in G_y$ ($G_x \in \groups, G_y \in \groups$) and each value in $g_y$ is the rank of the corresponding value in $g_x$ (possibly with ties).
\end{itemize}
When looking into constraints like $\ecrank{G_r}{G_x}$, it is helpful to see the analogy with both first order logic (FOL) and with constraint satisfaction problems.
From a FOL perspective, the name of the constraint (\textit{RANK}) is just the predicate, and the arguments $B$ and $A$ are the terms, which can be seen as either uninstantiated variables or as concrete values.

This also holds in our setting, where an instantiation of a variable corresponds to a concrete set of vectors (a subgroup). For example for $\ecrank{G_r}{G_x}$ and the spreadsheet in Figure~\ref{fig:main_example}, when we write $\ecrank{\range{T_1}{\rangeall}{8}}{\range{T_1}{\rangeall}{7}}$, then the value of $G_r$ is the $8$th vector in $T_1$: $G_r = \range{T_1}{\rangeall}{8} = \{[2,1,3,4]\}$ and the value of $G_x$ is the $7$th vector: $G_x = \range{T_1}{\rangeall}{7} = \{[1514, 1551, 691, 392]\}$.
\tias{not OK to use set because of column-wise sum? we must properly define it then...}

\newcommand{\sigc}{\ensuremath{\format{Sig}_c}}
\newcommand{\defc}{\ensuremath{\format{Def}_c}}

With this interpretation, we can speak about the signature and definition of a constraint template being satisfied. We say that a signature (definition) of a constraint template with $n$ arguments is satisfied by an assignment $(G_1, ..., G_n)$ if $\sigc(G_1, ..., G_n)$ (respectively $\defc(G_1, ..., G_n)$) is satisfied. Likewise, an assignment $(G_1, ..., G_n)$ satisfies a constraint template $s$ iff $|(G_1, ..., G_n)| = n_s \wedge \sigc(G_1, ..., G_n) \wedge \defc(G_1, ..., G_n)$.

@@

\sergey{under question now, might worth deleting} \\
\sergey{example new template: vr = floor(v), float(v), int(vr), len(v)=len(vr), v,vr are single vectors?}
